\pkg{RProtoBuf}: Efficient Cross-Language Data Serialization in R

### 2014-01-14 papers/jss/article.Rnw
Edits to section 1 suggested by Karl.
1) Remove the Hadley Split-Apply-Combine reference for now as it is
confusing and more narrowly R-only than the
multi-platform-data-analysis pipeline pattern otherwise being
discussed in the first paragraph.  We might be able to add it back
with suitable distinctions added.

2) Note that our goal is not just to safely transfer the data, but to
safely _and efficiently_ do so.

3) Rewrite the second to last paragraph a bit, and add a note
specifically at the end that this paper describes an R implementation
of protocol buffers.



git-svn-id: svn://svn.r-for

Modern data collection and analysis pipelines are increasingly being
built using collections of components to better manage software
complexity through reusability, modularity, and fault
isolation \citep{Wegiel:2010:CTT:1932682.1869479}.
Data analysis patterns such as Split-Apply-Combine
\citep{wickham2011split} explicitly break up large problems into
manageable pieces.  These patterns are frequently employed with
different programming languages used for the different phases of data
analysis -- collection, cleaning, modeling, analysis, post-processing, and
presentation in order to take advantage of the unique combination of
performance, speed of development, and library support offered by
different environments and languages.  Each stage of such a data
analysis pipeline may involve storing intermediate results in a
file or sending them over the network.

Given these requirements, how do we safely share intermediate results
between different applications, possibly written in different
languages, and possibly running on different computer system, possibly
spanning different operating systems?  Programming
languages such as R, Julia, Java, and Python include built-in
serialization support, but these formats are tied to the specific
% DE: need to define serialization?
programming language in use and thus lock the user into a single
environment.

@@

Once the data serialization needs of an application become complex
enough, developers typically benefit from the use of an
\emph{interface description language}, or \emph{IDL}.  IDLs like
Protocol Buffers \citep{protobuf}, Apache Thrift, and Apache Avro provide a compact
well-documented schema for cross-language data structures and
efficient binary interchange formats.  The schema can be used to
generate model classes for statically-typed programming languages such
as C++ and Java, or can be used with reflection for dynamically-typed
programming languages.  Since the schema is provided separately from
the encoded data, the data can be efficiently encoded to minimize
storage costs of the stored data when compared with simple
``schema-less'' binary interchange formats.
Many sources compare data serialization formats and show Protocol
Buffers very favorably to the alternatives; see
\citet{Sumaray:2012:CDS:2184751.2184810} for one such comparison.

@@

The rest of the paper is organized as follows. Section~\ref{sec:protobuf}
provides a general overview of Protocol Buffers.
Section~\ref{sec:rprotobuf-basic} describes the interactive R interface
provided by \CRANpkg{RProtoBuf} and introduces the two main abstractions:
\emph{Messages} and \emph{Descriptors}.  Section~\ref{sec:rprotobuf-classes}
describes the implementation details of the main S4 classes making up this
package.  Section~\ref{sec:types} describes the challenges of type coercion
between R and other languages.  Section~\ref{sec:evaluation} introduces a
general R language schema for serializing arbitrary R objects and evaluates
it against R's built-in serialization.  Sections~\ref{sec:opencpu}
and \ref{sec:mapreduce} provide real-world use cases of \CRANpkg{RProtoBuf}
in web service and MapReduce environments, respectively, before
Section~\ref{sec:summary} concludes.

>>>>

Modern data collection and analysis pipelines are increasingly being
built using collections of components to better manage software
complexity through reusability, modularity, and fault
isolation \citep{Wegiel:2010:CTT:1932682.1869479}.
% This is really a different pattern not connected well here.
%Data analysis patterns such as Split-Apply-Combine
%\citep{wickham2011split} explicitly break up large problems into manageable pieces.  
These pipelines are frequently built with
different programming languages used for the different phases of data
analysis -- collection, cleaning, modeling, analysis, post-processing, and
presentation in order to take advantage of the unique combination of
performance, speed of development, and library support offered by
different environments and languages.  Each stage of such a data
analysis pipeline may involve storing intermediate results in a
file or sending them over the network.

Given these requirements, how do we safely and efficiently share intermediate results
between different applications, possibly written in different
languages, and possibly running on different computer system, possibly
spanning different operating systems?  Programming
languages such as R, Julia, Java, and Python include built-in
serialization support, but these formats are tied to the specific
% DE: need to define serialization?
programming language in use and thus lock the user into a single
environment.

@@

Once the data serialization needs of an application become complex
enough, developers typically benefit from the use of an
\emph{interface description language}, or \emph{IDL}.  IDLs like
Protocol Buffers \citep{protobuf}, Apache Thrift, and Apache Avro provide a compact
well-documented schema for cross-language data structures and
efficient binary interchange formats.
Since the schema is provided separately from the encoded data, the data can be
efficiently encoded to minimize storage costs of the stored data when compared with simple
``schema-less'' binary interchange formats. Many sources compare data serialization formats
and show Protocol Buffers compare very favorably to the alternatives; see
\citet{Sumaray:2012:CDS:2184751.2184810} for one such comparison.
The schema can be used to generate classes for statically-typed programming languages
such as C++ and Java, or can be used with reflection for dynamically-typed programming
languages

%  The schema can be used to
%generate model classes for statically-typed programming languages such
%as C++ and Java, or can be used with reflection for dynamically-typed
%programming languages.  Since the schema is provided separately from
%the encoded data, the data can be efficiently encoded to minimize
%storage costs of the stored data when compared with simple
%``schema-less'' binary interchange formats.
%Many sources compare data serialization formats and show Protocol
%Buffers very

@@

This paper describes an R interface to protocol buffers.
The rest of the paper is organized as follows. Section~\ref{sec:protobuf}
provides a general overview of Protocol Buffers.
Section~\ref{sec:rprotobuf-basic} describes the interactive R interface
provided by \CRANpkg{RProtoBuf} and introduces the two main abstractions:
\emph{Messages} and \emph{Descriptors}.  Section~\ref{sec:rprotobuf-classes}
describes the implementation details of the main S4 classes making up this
package.  Section~\ref{sec:types} describes the challenges of type coercion
between R and other languages.  Section~\ref{sec:evaluation} introduces a
general R language schema for serializing arbitrary R objects and evaluates
it against R's built-in serialization.  Sections~\ref{sec:opencpu}
and \ref{sec:mapreduce} provide real-world use cases of \CRANpkg{RProtoBuf}
in web service and MapReduce environments, respectively, before
Section~\ref{sec:summary} concludes.
